---
title: "PSYC234_Lab_2_Student"
author: "Emma Mills"
date: "23/01/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
# library(gdata)
library(PerformanceAnalytics)
library(effects)
library(gridExtra)
```


# Multiple Regression with Categorical Predictors

Many factors may predict academic success in young people that are not part of individual differences - i.e. their ability for a subject.  In this session, we are going to predict final Portuguese test scores (G3) from seven factors:

 - sex 
 - their health
 - their school
 - their school attendance rate
 - travel time to school
 - whether they attend extra private tutor classes
 - how much they drink at the weekend
 
Use the `Data_Codebook.pdf` file in the lab folder to find out initial information such as variable names, their type and their levels or numerical ranges.


## Import Data

### Portuguese Scores

```{r}
d_raw <- read_csv("student-por.csv") # Portuguese scores

```

## Tidying

You could choose to reduce the dataset to keep only the variables they need

Select out the columns for manageability:  use the select() function

```{r}

# select the columns you need and put them in any order you want.

# keep the raw data set separate, because you might want to describe the sample and will need age etc. for this.

```

## Visualise to get to know your data:

### Predictors

There is code above  to help you do this for the categorical and continuous predictors.

Prompt questions to interrogate what you see:

 - Are the categorical variables balanced?
 - Are they made of two levels or more?
 - Do any of the categorical variables look like they might be ordinal?
 - How do the x-axes of the continuous variables look - does the range seem sensible?
 - Is there any skew to the continuous variables.
 
### Outcome variable

  - Is it a continuous variable?
  - Or if not, can we treat it as such for our test?

```{r message=FALSE, warning=FALSE}
# bar charts for categorical variables



# (so called) continuous variables


grid.arrange(p_sex, p_sch, p_paid, p_health, p_absence, p_travel, p_walc, p_outcome)
```

Write yourself some text here as descriptors or what you see that helps you to get to know this unfamiliar dataset.


### Correlations betweeen continuous predictors

In the code below,  there is a bracket `c( )`.  You need to type the number of the columns that represent the columns in your dataset - so what number column is the outcome predictor etc.  Separate each number with a comma.

```{r warning=FALSE}
chart.Correlation(d[, c( )], histogram=TRUE, pch=19)
```

Prompt questions to interrogate what you see:

thresholds to describe the relationships: 0.1 - 0.3 is small; 0.3 - 0.5 is medium; 0.5 - 0.7 is large.

  - What size of correlation do you see between the variables?
  - Do the correlations between predictors and the outcome variable suggest what you might see in the regression model?
  
Remember that these correlations are *bivariate* correlations - explaining shared variance between only those two predictors.  When we put the predictors into a regression model, we are estimating the independent variance of a predictor AFTER we have explained some of the variance already - in the other predictors. 

## Model

### Raw Variables

Write a model and call a summary of the model

```{r}

```
## Transforming Categorical Variables

Given the predictors, and having watched the lecture, what is the best decision you can make about how the catgorical variables are coded in the model?

Consider:

 - dummy coding scheme
 - sum coding scheme
 - transforming the catgorical variable into a numerical variable and centring
 - helmert coding
 
Make your decisions according to what is suggested by the model output and perform the transformations.  

If you want to use your script for revision,  it is worthwhile noting on the script why you are making the decisions that you are making - that way, if you submit for feedback, I can check your reasoning.

## Transforming Continuous Variables to Work with Categorical Variables

Consider:

 - raw variables
 - centred variables
 - standardised variables
 
Make your decisions according to what is suggested by the model output and perform the transformations.  

Here is a blank r chunk that you can copy and paste to write your commands if you are working in RMarkdown.
The keyboard shortcut to do that is Alt-Shift-I
The menu item is Code >> Insert Chunk

```{r}
```



```{r}

```


## Model 


```{r}

```

## Communicate

### Descriptive Statistics

You should report how many females / males,  the mean age of both sexes and how the sample was split by school.

### Report the Model

You should report:

  - the name of the outcome variable
  - the names of the predictors in the model
  - then the omnibus test for the model (the F ratio and associated information)
  - then the results of each predictor - i.e. report the levels of walc together.
  - when you are reporting the statistical results of each predictor, make sure you give a verbal description of the result when the coefficient is significant.
  
Draw a table - remember to refer to the Table in your textual description.

### Plotting Predictions

You can communicate through plots also.  Plot some predictor effects that you could use in a results section. 

Use the screenshots of predictor effects as a self-check

Remember to refer to them as Figures in your textual descriptions.

```{r}


```



