---
title: "Multiple_Regression_Mediation_Student"
output:
  html_document: default
  word_document: default
  
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(include = TRUE)
library(tidyverse)
library(effects)
library(broom)
library(gridExtra)
library(PerformanceAnalytics)
library(diagram)
library(mediation)
```

### Mediation analysis

In this demonstration, we will model a mediation analysis. The total effect of an unmediated relationship is below in pathway c.

```{r}
data <- c(0, "a", 0,
          0, 0, 0, 
          "b", "c", 0)
M<- matrix (nrow=3, ncol=3, byrow = TRUE, data=data)
plot<- plotmat (M, pos=c(1,2), 
                name= c( "M",
                         "Coffee (X)", 
                         "Problem \nSolving (Y)"), 
                box.type = "rect", 
                box.size = 0.12, 
                box.prop=0.5,  
                curve=0)

```

We are going to test for the mediator variable of Alertness (M) on Coffee's (X) performance on Problem Solving (Y).  (This needs the prime mark added to the direct pathway "c" text!)

```{r}
data <- c(0, "a", 0,
          0, 0, 0, 
          "b", "`c (direct effect)`", 0)
M<- matrix (nrow=3, ncol=3, byrow = TRUE, data=data)
plot<- plotmat (M, pos=c(1,2), 
                name= c( "Alertness (M)",
                         "Coffee (X)", 
                         "Problem \nSolving (Y)"), 
                box.type = "rect", 
                box.size = 0.12, 
                box.prop=0.5,  
                curve=0)

```

### Read in the data

```{r}
library(readr)
d <- read_csv("Coffee_Mediation_1.csv")

head(d)
```


```{r}
summary(d)
```

I am going to copy and rename the variables to save on typing

X = Coffee,
Y = Prob_Solve,
M = Alert
 
```{r}
d <- d %>% 
  mutate(X = Coffee,
         Y = Prob_Solve,
         M = Alert)
```
 

### Longhand - Steps of Baron & Kenny (1986)

Four independent linear regression models

  1. The effect of X on Y (total effect = pathway c) = a simple linear regression model
  2. The effect of X on M (indirect effect pathway a) = a simple linear regression model
  3. The effect of M on Y (indirect effect pathway b) while controlling for X = a multiple linear regression model
  4. The effect of X and M on Y (for direct effect estimation = pathway c') = the same multiple linear regression model
  
When running your models, you need to assign them to objects in the environment to then be able to use them in a call to the `mediation` package.

Run the code for the mediation analysis and interpret the results at each step.

#### Step 1: Test the total effect - pathway c

$$
Y = b_0 + b_1 * X + e
$$

#### Construct a simple regression model for X predicting Y
```{r total effect}
(mod_totaleffect <-  summary(lm(Y ~ X, d)))
```

Write a short paragraph describing the model findings (about 2 or three sentences):

 - Describe the relationship between X and Y in terms of significance
 - Describe the magnitude of the effect of X on Y



#### Step 2: Simple Regression Model to test the a pathway of the indirect effect

$$
M = b_0 + b_1 * X + e
$$
#### Construct a simple regression model for X predicting M
```{r indirect effect a}
(mod_a <- summary(lm(M ~ X, d)))
```

Write a short paragraph describing the model findings (about 2 or three sentences):

 - Describe the relationship between X and M in terms of significance
 - Describe the magnitude of the effect of X on M
 - Does X predict M - include a sentence that indicates whether or not the relationship between X and M allows for M to act as a mediator, given the additional assumptions about mediation analysis.



#### Step 3 & 4: Construct a multiple regression to test the b pathway of the indirect effect & the direct effect (pathway c')

$$
Y = b_0 + b_1 * X + b_2 * M + e
$$

```{r indirect effect b}
(mod_b <- summary(lm(Y ~ X + M, d)))
```

Write a short paragraph describing the model findings (about three or four sentences):

 - Describe the relationship between M and Y in terms of significance
 - Describe the magnitude of the effect of M on Y
 - Describe the new relationship of X on Y and compare it to the model of total effect
 - Describe whether and what type of mediation effect is present between the three variables


### Using the power of R and the `mediation` package

The `mediation` package is loaded at the top of the document.  It takes the models for pathways a and b, and it needs us to tell it the name of the predictor or treatment variable (X) and the name of the mediator variable (M) and we set the `boot` argument to T for TRUE, to be able to generate confidence intervals on our co-efficients.  Refer to the lecture model for code if you are unsure.

```{r}
results <- mediate(mod_a, mod_b, treat = 'X', mediator = 'M', boot = T, dropobs = TRUE)
```

run a summary of the results:

```{r}
summary(results)
```

 - **ACME** stands for *average causal mediation effects* and is the product of pathway a  (X = ) and pathway b (M = ).
 - **ADE** stands for average direct effects or pathway c'.  This is the X coefficient in our pathway b model
 - **Total Effect** does what it says on the tin.  It is the sum of the direct and indirect effect, ACME + ADE, and also calculated  as X in our total effect model
 - **Prop. Mediated** is the proportion of the effect of X on Y that goes through M.  We divide ACME (or ab) by the total effect (c).
 

Drawing the diagram from the top of the document:

In the chunk below, follow the instructions to change a part of the code to reflect our mediation relationship.

In the command to data,  

 - remove the quotation marks, back ticks and question marks,
 - put in the correct coefficients to reflect the indirect a, indirect b and direct pathways,
 - make sure that you have a comma after each changed part of code

```{r}
data <- c(0, "`?`", 0,
          0, 0, 0, 
          "`?`", "`?`", 0)
M <- matrix (nrow=3, ncol=3, byrow = TRUE, data=data)
plot<- plotmat (M, pos=c(1,2), 
                name= c( "Alert (M)",
                         "Coffee (X)", 
                         "Problem \nSolving (Y)"), 
                box.type = "rect", 
                box.size = 0.12, 
                box.prop=0.5,  
                curve=0)

```

### Reporting the mediation analysis

Write a results paragraph reporting the mediation effect:  Use the lecture example as a model.  It should describe the mediation effect and also report your use of a bootstrapping procedure.






How are we doing with our list now, with one half of a lecture to go!!

  - a model with some significant predictors! - done
  - correlation matrices - next week
  - centering predictors - done
  - standardising predictors - done
  - models with categorical predictors - done & ongoing
    - interpreting models with categorical predictors - done & ongoing
  - models with interaction terms - done & ongoing
  - the properties of multicollinearity - next week
  - choosing between different models - done
  - reporting models - ongoing